---
title: Accessing Gen3 in the AnVIL
author:
- name: BJ Stubbs
- name: Martin Morgan
package: Gen3
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Accessing Gen3 in the AnVIL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
cache <- tools::R_user_dir("Gen3", "cache")
credentials <- file.path(cache, "credentials.json")
knitr::opts_chunk$set(eval = file.exists(credentials))
```

# Introduction

This is a demo of how to login and query Gen3 from R.

This is based on:

* https://gen3.org/resources/user/using-api/
* https://gen3.org/resources/developer/flat-model-api/
* https://graphql.org/learn/queries/
* https://graphql.org/learn/introspection/

A very useful facility for formulating GraphQL queries is

* https://gen3.theanvil.io/query

## Package installation and loading

If necessary, install the Gen3 library

```{r}
if (!"Gen3" %in% rownames(installed.packages()))
    BiocManager::install("Bioconductor/Gen3")
```

Load the library into the current _R_ session. Also useful for this
vignette is the dplyr package.

```{r}
library(Gen3)
library(dplyr)
```

## Authentication

Go to this url: https://gen3.theanvil.io and login, then click on the
profile link. There you can create an access credential. Download this
json file and remember its location.

Load the access token from the credentials file and authenticate with
Gen3. Adjust the value of `credentials` to reflect the location of the
credentials file you obtained from Gen3.

```{r}
cache <- tools::R_user_dir("Gen3", "cache")
credentials <- file.path(cache, "credentials.json")
authenticate(credentials)
```

If a session has been idle for a while, the authentication credentials
may expire, resulting in a message like

```
Error in .query(body) : Unauthorized (HTTP 401).
```

Simply renew credentials using `authenticate()` as described above.

# Schema discovery and principles of data exploration

Get schema types for Gen3. These contain the root entities for queries.

```{r}
## { __schema { types { name } } }
schema()
```

`schema("full")` provides a more complete list of schema entities. The
GraphQL query performed by this function is summariized in the
comment.

Each schema entry is associated with fields; discover these with, e.g.,

```{r}
## { __type(name: subject) { fields{ name } } }
fields("subject")
```

`values()` performs a query against the database. The first argument
is the name of the entity to be retrieved; subsequent arguments are
the fields within that entity. The number of records returned is given
by `.n`, which has a default value of 10. Use `.n = 0` to retrieve all
entities.

```{r}
## { subject(first: 50) { id project_id sex } }
values("subject", "id", "project_id", "sex", .n = 50)
```

# An initial exploration

## Projects

The first query we will do is to find all of the projects that we have
access to. `projects()` returns the fields `project_id`, `id`, and
`study_description` from all projects we have access to.

```{r}
## { project(first: 0) { project_id id study_description } }
projects()
```

## Subject, sample and sequencing entities

The main entities in Gen3 are subject, sample, and sequencing. Get
the fields available in the subject entity like this:

```{r}
## { __type(name: subject) { fields { name } } }
fields("subject") # any `name` of schema()
```

Similarly, the fields for 'sample' and 'sequencing' are

```{r}
## { __type(name: sample) { fields { name } } }
fields("sample")

## { __type(name: sequencing) { fields { name } } }
fields("sequencing")
```

Query the value of fields, across all projects, with

```{r}
## { sample(first: 10) { id rin_number } }
values("sample", "id", "rin_number", .n = 10)

## { sequencing(first: 10) { id file_name } }
values("sequencing", "id", "file_name")
```

The Gen3 schema attempts to represent subjects (for example) from all
studies in a single entity, so the value of many fields may be missing
(`null` in the GraphQL respond, represented as `NA` in _R_).

```{r}
values("subject", "project_id", "weight", "age_of_onset")
```

# Direct use of GraphQL

Use GraphQL directly for more complicated queries, e.g., filtering parameters.

For example queries the `subject` field restricted to those with the
"open_access-1000Genomes" `project_id` for four values: `id`, `sex`,
`population`, and `submitter_id`. Using `first: 0` returns all
records.

```{r}
query <- '{
    subject(
        project_id: "open_access-1000Genomes"
        first: 0
    ) {
        id
        sex
        population
        submitter_id
    }
}'
result <- query_graphql(query)
```

The return value is a list with structure like that of the
query. values for each subject have been simiplified to a tibble. The
elements of the list are accessible using standard _R_ operations

```{r}
result$subject
```

The following query retrieves the `id` and `tissue_type` of all samples in GTEx.

```{r}
query <- '{
    sample(
        project_id: "CF-GTEx"
        first: 0
    ) {
        id
        tissue_type
    }
}'
result <- query_graphql(query)
result$sample
```

The tibble is easily explored using standard tidy paradigms, e.g.,

```{r}
result$sample %>%
    count(tissue_type) %>%
    arrange(desc(n))
```

Syntax errors return an error from the server

```{r}
query <- '{ sample { id }'
tryCatch(query_graphql(query), error = identity)
```

# Session information

```{r}
sessionInfo()
```
